### 2019-09-29 ch6_bred/content.tex
clean up thesis

\section{Introduction}
Single-cell omics data, including transcriptomics, proteomics and epigenomics data, provide new opportunities for studying cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation \cite{tanay_scalingsinglecellgenomics_2017,etzrodt_quantitativesinglecellapproaches_2014}. Such dynamic processes can be modeled computationally using trajectory inference (TI) methods, also called pseudotime analysis, which order cells along a trajectory based on similarities in their expression patterns \cite{trapnell_definingcelltypes_2015,cannoodt_computationalmethodstrajectory_2016,moon_manifoldlearningbasedmethods_2018}. The resulting trajectories are most often linear, bifurcating or tree-shaped, but more recent methods also identify more complex trajectory topologies, such as cyclic \cite{liu_reconstructingcellcycle_2017} or disconnected graphs \cite{wolf_graphabstractionreconciles_2017}. TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process\cite{tanay_scalingsinglecellgenomics_2017}, thereby allowing the objective identification of new (primed) subsets of cells \cite{schlitzer_identificationcdc1cdc2committed_2015}, delineation of a differentiation tree \cite{velten_humanhaematopoieticstem_2017,see_mappinghumandc_2017} and inference of regulatory interactions responsible for one or more bifurcations \cite{aibar_scenicsinglecellregulatory_2017}. Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogs of whole organisms \cite{regev_humancellatlas_2017,han_mappingmousecell_2018,schaum_singlecelltranscriptomics20_2018} underline the urgency for accurate, scalable \cite{aibar_scenicsinglecellregulatory_2017,angerer_singlecellsmake_2017} and user-friendly TI methods.

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table 1). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

Given the diversity in TI methods, it is important to quantitatively assess their performance, scalability, robustness and usability. Many attempts at tackling this issue have already been made \cite{haghverdi_diffusionpseudotimerobustly_2016,ji_tscanpseudotimereconstruction_2016,welch_slicerinferringbranched_2016,matsumoto_scoupprobabilisticmodel_2016,duverle_celltreebioconductorpackage_2016,cannoodt_computationalmethodstrajectory_2016,lonnberg_singlecellrnaseqcomputational_2017,campbell_probabilisticmodelingbifurcations_2017,wolf_graphabstractionreconciles_2017}, but a comprehensive comparison of TI methods across a large number of different datasets is still lacking. This is problematic, as new users to the field are confronted with an overwhelming choice of TI methods, without a clear idea of which would optimally solve their problem. Moreover, the strengths and weaknesses of existing methods need to be assessed, so that new developments in the field can focus on improving the current state-of-the-art.

In this study, we evaluated the accuracy, scalability, stability and usability of 45 TI methods (Figure \ref{fig:figure_1}a). We found substantial complementarity between current methods, with different sets of methods performing most optimally depending on the characteristics of the data. For method users, we created an interactive set of guidelines (available at \href{http://guidelines.dynverse.org}{guidelines.dynverse.org}), which gives context-specific recommendations for method usage. Our evaluation also highlights some challenges for current methods, and our evaluation strategy can be useful to spearhead the development of new tools that accurately infer trajectories on ever more complex use cases.

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_1.pdf}
	\caption{
		\textbf{Overview of several key aspects of the evaluation.}
		\textbf{a}, A schematic overview of our evaluation pipeline. \textbf{b}, To make the trajectories comparable to each other, a common trajectory model was used to represent reference trajectories from the real and synthetic datasets, as well as any predictions of TI methods. \textbf{c}, Trajectories are automatically classified into one of seven trajectory types, with increasing complexity. \textbf{d}, We defined four metrics, each assessing the quality of a different aspect of the trajectory. The HIM score assesses the similarity between the two topologies, taking into account differences in edge lengths and degree distributions. The F1branches assesses the similarity of the assignment of cells onto branches. The cordist quantifies the similarity in cellular positions between two trajectories, by calculating the correlation between pairwise geodesic distances. Finally, wcorfeatures quantifies the agreement between trajectory differentially expressed features from the known trajectory and the predicted trajectory.
	}
	\label{fig:figure_1}
\end{figure}

\section{Results}

\subsection{Trajectory inference methods}
To make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (Figure \ref{fig:figure_1}b). In this model, the overall topology is represented by a network of ‘milestones’, and the cells are placed within the space formed by each set of connected milestones. Although almost every method returned a unique set of outputs, we were able to classify these outputs into seven distinct groups (Figure \ref{fig:supfigure_1}) and we wrote a common output converter for each of these groups (Figure \ref{fig:figure_2}a). When strictly required, we also provided prior information to the method. These different priors can range from weak priors that are relatively easy to acquire, such as a start cell, to strong priors, such as a known grouping of cells, that are much harder to know a priori, and which can potentially introduce a large bias into the analysis (Figure \ref{fig:figure_2}a).

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/supfigure_1.pdf}
	\caption{
		\textbf{A common interface for TI methods.}
		\textbf{a} The input and output of each TI method is standardized. As input, each TI method receives either raw or normalized counts, several parameters, and a selection of prior information. After its execution, a method uses one of the seven wrapper functions to transform its output to the common trajectory model. This common model then allows to perform common analysis functions on trajectory models produced by any TI method. \textbf{b} Illustrations of the specific transformations performed by each of the wrapper functions.
	}
	\label{fig:supfigure_1}
\end{figure}

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_2.pdf}
	\caption{
		\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
		\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig. 2.
	}
	\label{fig:figure_2}
\end{figure}

The largest difference between TI methods is whether a method fixes the topology and, if it does not, what kind of topology it can detect. We defined seven possible types of topology, ranging from very basic topologies (linear, cyclical and bifurcating) to the more complex ones (connected and disconnected graphs). Most methods either focus on inferring linear trajectories or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (Figure \ref{fig:figure_2}a).

We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b). We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig. 2), after which we conclude with guidelines for method users and future perspectives for method developers.

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_3.pdf}
	\caption{	
		\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
		\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
	}
	\label{fig:figure_3}
\end{figure}

\subsection{Accuracy}

We defined several metrics to compare a prediction to a reference trajectory (Supplementary Note 1). Based on an analysis of their robustness and conformity to a set of rules (Supplementary Note 1), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table 2). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

We found that method performance was very variable across datasets, indicating that there is no ‘one-size-fits-all’ method that works well on every dataset (Figure \ref{fig:supfigure_3}a). Even methods that can detect most of the trajectory types, such as PAGA, RaceID/StemID and SLICER were not the best methods across all trajectory types (Figure \ref{fig:figure_3}b). The overall score between the different dataset sources was moderately to highly correlated (Spearman rank correlation between 0.5–0.9) with the scores on real datasets containing a gold standard (Figure \ref{fig:supfigure_3}b), confirming both the accuracy of the gold standard trajectories and the relevance of the synthetic data. On the other hand, the different metrics frequently disagreed with each other, with Monocle and PAGA Tree scoring better on the topology scores, whereas other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (Figure \ref{fig:figure_3}b).

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/supfigure_3.pdf}
	\caption{
		\textbf{Accuracy of trajectory inference methods.}
		\textbf{a} Overall score for all methods across 339 datasets, colored by the source of the datasets. Black line indicates the mean. \textbf{b} Similarity between the overall scores of all dataset sources, compared to real datasets with a gold standard, across all methods (n = 46, after filtering out methods that errored too frequently). Shown in the top left is the Pearson correlation. \textbf{c} Bias in the overall score towards trajectory types for all methods across 339 datasets. Black line indicates the mean. \textbf{d} Distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.
	}
	\label{fig:supfigure_3}
\end{figure}

The performance of a method was strongly dependent on the type of trajectory present in the data (Figure \ref{fig:figure_3}b). Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories (Figure \ref{fig:supfigure_3}c). This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, whereas those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (Figure \ref{fig:supfigure_3}d). This analysis therefore indicates that detecting the right topology is still a difficult task for most of these methods, because methods tend to be either too optimistic or too pessimistic regarding the complexity of the topology in the data.

The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we calculated the likelihood of obtaining a top model when using only a subset of all methods. A top model in this case was defined as a model with an overall score of at least 95$\%$ as the best model. On all datasets, using one method resulted in getting a top model about 27$\%$ of the time. This increased up to 74$\%$ with the addition of six other methods (Figure \ref{fig:figure_4}a). The result was a relatively diverse set of methods, containing both strictly linear or cyclic methods, and methods with a broad trajectory type range such as PAGA. We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (Figure \ref{fig:figure_4}b), although in these cases less methods were necessary to obtain at least one top model for a given dataset. Altogether, this shows that there is considerable complementarity between the different methods and that users should try out a diverse set of methods on their data, especially when the topology is unclear a priori. Moreover, it also opens up the possibilities for new ensemble methods that utilize this complementarity.

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_4.pdf}
	\caption{
		\textbf{Complementarity between different trajectory inference methods.}
		\textbf{a}, We assessed the likelihood for different combinations of methods to lead to a ‘top model’ (defined as a model with an overall score of at least 95$\%$ of the best model) when applied to all datasets. \textbf{b}, The likelihood for different combinations of methods to lead to a ‘top model’ was assessed separately on different trajectory types. For this figure, we did not include any methods requiring a cell grouping or a time course as prior information.
	}
	\label{fig:figure_4}
\end{figure}

\subsection{Scalability}

While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a). As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Supplementary Fig. 5), and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig. 2a).

\begin{figure}[htb!]
	\centering\includegraphics[width=0.9\linewidth]{fig/supfigure_4.pdf}
	\caption{
		\textbf{Scalability of trajectory inference methods.}
		\textbf{a} Three examples of average observed running times across five datasets (left) and the predicted running time (right). \textbf{b} Overview of the scalability results of all methods, ordered by their average predicted running time from (a). We predicted execution times and memory usage for each method with increasing number of features or cells, and used these values to classify each method into sublinear, linear, quadratic and superquadratic based on the shape of the curve.
	}
	\label{fig:supfigure_4}
\end{figure}

We found that the scalability of most methods was overall very poor, with most graph and tree methods not finishing within an hour on a dataset with ten thousand cells and ten thousand features (Figure \ref{fig:figure_3}c), which is around the size of a typical droplet-based single-cell dataset \cite{svensson_exponentialscalingsinglecell_2018}. Running times increased further with increasing number of cells, with only a handful of graph/tree methods completing within a day on a million cells (PAGA, PAGA Tree, Monocle DDRTree, Stemnet and GrandPrix). Some methods, such as Monocle DDRTree and GrandPrix, also suffered from unsatisfactory running times when given a high number of features.

Methods with a low running time typically had two defining aspects: they had a linear time complexity with respect to the features and/or cells, and adding new cells or features led to a relatively low increase in time (Figure \ref{fig:supfigure_4}b). We found that more than half of all methods had a quadratic or superquadratic complexity with respect to the number of cells, which would make it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand cells (Figure \ref{fig:supfigure_4}b).

We also assessed the memory requirements of each method (Supplementary Fig. 2c). Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features. Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).

Altogether, the scalability analysis indicated that the dimensions of the data are an important factor in the choice of method, and that method development should pay more attention to maintaining reasonable running times and memory usage.
§
\subsection{Stability}

It is not only important that a method is able to infer an accurate model in a reasonable time frame, but also that it produces a similar model when given very similar input data. To test the stability of each method, we executed each method on ten different subsamples of the datasets (95$\%$ of the cells, 95$\%$ of the features), and calculated the average similarity between each pair of models using the same scores used to assess the accuracy of a trajectory (Figure \ref{fig:figure_3}d).

Given that the trajectories of methods that fix the topology either algorithmically or through a parameter are already very constrained, it is to be expected that such methods tend to generate very stable results. Nonetheless, some fixed topology methods still produced slightly more stable results, such as SCORPIUS and MATCHER for linear methods and MFA for multifurcating methods. Stability was much more diverse among methods with a free topology. Slingshot produced more stable models than PAGA (Tree), which in turn produced more stable results than pCreode and Monocle DDRTree.

\subsection{Usability}

While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_tensimplerules_2017}. We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}). It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB). The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable). Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.

We found that most methods fulfilled the basic criteria, such as the availability of a tutorial and elemental code quality criteria (Figure \ref{fig:figure_3}d and Supplementary Fig. 6). While recent methods had a slightly better quality score than older methods, several quality aspects were consistently lacking for the majority of the methods (Supplementary Fig. 6 right) and we believe that these should receive extra attention from developers. Although these outstanding issues covered all five categories, code assurance and documentation in particular were problematic areas, notwithstanding several studies pinpointing these as good practices \cite{wilson_bestpracticesscientific_2014,artaza_top10metrics_2016}. Only two methods had a nearly perfect usability score (Slingshot and Celltrails), and these could be used as an inspiration for future methods. We observed no clear relation between usability and method accuracy or usability (Figure \ref{fig:figure_2}b).

\begin{table}
	\caption{\textbf{Scoring sheet for assessing usability of trajectory inference methods.} Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development.} \label{tab:scoresheet}

	\centering\begingroup\fontsize{7}{9}\selectfont

	\begin{tabular}{p{3.5cm}p{9cm}p{2.5cm}}
		\toprule
		Aspect & Items & References \\ \midrule

		\multicolumn{3}{l}{\textbf{Availability}}\\*

		Open source & (1) Method's code is freely available (2) The code can be run on a freely available platform & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

		Version control & The code is available on a public version controlled repository, such as Github & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*

		Packaging & (1) The code is provided as a "package", exposing functionality through functions or shell commands (2) The code can be easily installed through a repository such as CRAN, Bioconductor, PyPI, CPAN, debian packages, … & \cite{lee_rpackagespackagedevelopment_2017,wickham_packagesorganizetest_2015,jimenez_foursimplerecommendations_2017,silva_generalguidelinesbiomedical_2017}\\*

		Dependencies & (1) Dependencies are clearly stated in the tutorial or in the code (2) Dependencies are automatically installed & \cite{taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,karimzadeh_topconsiderationscreating_}\\*

		License & (1) The code is licensed (2) License allows academic use & \cite{lee_rpackagespackagedevelopment_2017,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

		Interface & (1) The tool can be run using a graphical user interface, either locally or on a web server (2) The tool can be run through the command line or through a programming language & \cite{silva_generalguidelinesbiomedical_2017}\\

		\midrule
		\multicolumn{3}{l}{\textbf{Code quality}} \\*

		Function and object naming & (1) Functions/commands have well chosen names (2) Arguments/parameters have well chosen names & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\*

		Code style & (1) Code has a consistent style (2) Code follows (basic) good practices in the programming language of choice, for example PEP8 or the tidyverse style guide & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016}\\*

		Code duplication & Duplicated code is minimal & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015}\\

		Self-contained functions & The method is exposed to the user as self-contained functions or commands & \cite{anderson_writinggreatscientific_2016,taschuk_tensimplerules_2017,silva_generalguidelinesbiomedical_2017}\\*

		Plotting & Plotting functions are provided for the final and/or intermediate results & \\*

		Dummy proofing & Package contains dummy proofing, i.e. testing whether the parameters and data supplied by the user make sense and are useful & \cite{lee_rpackagespackagedevelopment_2017,karimzadeh_topconsiderationscreating_}\\

		\midrule
		\multicolumn{3}{l}{\textbf{Code assurance}} \\*

		Unit testing & Method is tested using unit tests & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*

		Unit testing & Tests are run automatically using functionality from the programming language & \cite{lee_rpackagespackagedevelopment_2017,wilson_bestpracticesscientific_2014,anderson_writinggreatscientific_2016,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017}\\*

		Continuous integration & The method uses continuous integration, for example on Travis CI & \cite{beaulieu-jones_reproducibilitycomputationalworkflows_2017,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017}\\*

		Code coverage & (1) The code coverage of the repository is assessed. (2) What is the percentage of code coverage & \\

		\midrule
		\multicolumn{3}{l}{\textbf{Documentation}} \\*

		Support & (1) There is a support ticket system, for example on Github (2) The authors respond to tickets and issues are resolved within a reasonable time frame & \cite{wilson_bestpracticesscientific_2014,wickham_packagesorganizetest_2015,artaza_top10metrics_2016,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017}\\*

		Development model & (1) The repository separates the development code from master code, for example using git master en developer branches (2) The repository has created releases, or several branches corresponding to major releases. (3) The repository has branches for the development of separate features. & \cite{driessen_successfulgitbranching_2010}\\*

		Tutorial & (1) A tutorial or vignette is available (2) The tutorial has example results (3) The tutorial has real example data (4) The tutorial showcases the method on several datasets (1=0, 2=0.5, >2=1) & \cite{wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,jimenez_foursimplerecommendations_2017,karimzadeh_topconsiderationscreating_,boulesteix_tensimplerules_2015}\\*

		Function documentation & (1) The purpose and usage of functions/commands is documented (2) The parameters of functions/commands are documented (3) The output of functions/commands is documented & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*

		Inline documentation & Inline documentation is present in the code & \cite{wilson_bestpracticesscientific_2014,taschuk_tensimplerules_2017,wickham_packagesorganizetest_2015,silva_generalguidelinesbiomedical_2017,karimzadeh_topconsiderationscreating_}\\*

		Parameter transparency & All important parameters are exposed to the user & \cite{taschuk_tensimplerules_2017}\\

		\midrule
		\multicolumn{3}{l}{\textbf{Behaviour}} \\*

		Seed setting & The method does not artificially become deterministic, for example by setting some (0.5) or a lot (1) of seeds & \cite{puget_greendiceare_2016}\\*

		Unexpected output & (1) No unexpected output messages are generated by the method (2) No unexpected files, folders or plots are generated (3) No unexpected warnings during runtime or compilation are generated & \cite{artaza_top10metrics_2016}\\*

		Trajectory format & The postprocessing necessary to extract the relevant output from the method is minimal (1), moderate (0.5) or extensive (0) & \\*

		Prior information & Prior information is required (0), optional (1) or not required (1) & \\*

		\midrule
		\multicolumn{3}{l}{\textbf{Paper}} \\*

		Publishing & The method is published & \\*

		Peer review & The paper is published in a peer-reviewed journal & \cite{karimzadeh_topconsiderationscreating_,gannon_essentialrolepeer_2001,baldwin_refereeswetrust_2017}\\*

		Evaluation on real data & (1) The paper shows the method's usefulness on several (1), one (0.25) or no real datasets. (2) The paper quantifies the accuracy of the method given a gold or silver standard trajectory & \cite{aniba_issuesbioinformaticsbenchmarking_2010,jelizarow_overoptimismbioinformaticsillustration_2010}\\*

		Evaluation of robustness & The paper assessed method robustness (to eg. noise, subsampling, parameter changes, stability) in one (0.5) or several (1) ways & \cite{karimzadeh_topconsiderationscreating_,aniba_issuesbioinformaticsbenchmarking_2010,boulesteix_tensimplerules_2015,jelizarow_overoptimismbioinformaticsillustration_2010}\\

		\bottomrule
	\end{tabular}
	\endgroup{}
\end{table}

\section{Discussion}

In this study, we presented a large-scale evaluation of the performance of 45 TI methods. By using a common trajectory representation and four metrics to compare the methods’ outputs, we were able to assess the accuracy of the methods on more than 200 datasets. We also assessed several other important quality measures, such as the quality of the method’s implementation, the scalability to hundreds of thousands of cells and the stability of the output on small variations of the datasets.

Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and guidelines.dynverse.org). We postulate that, as a method’s performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data. For most use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, our evaluation suggests a different set of optimal methods, as shown in Figure \ref{fig:figure_5}. Several other factors will also impact the choice of methods, such as the dimensions of the dataset and the prior information that is available. These factors and several others can all be dynamically explored in our interactive app (guidelines.dynverse.org). This app can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_5.pdf}
	\caption{
		\textbf{Practical guidelines for method users.}
		As the performance of a method mostly depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user’s existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method’s performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the accuracy (+: scaled performance $\geq$ 0.9, $\pm$: >0.6), usability scores (+:$\geq$0.9, $\pm$ $\geq$0.6), estimated running times and required prior information. k, thousands; m, millions.
	}
	\label{fig:figure_5}
\end{figure}

When inferring a trajectory on a dataset of interest, it is important to take two further points into account. First, it is critical that a trajectory, and the downstream results and/or hypotheses originating from it, are confirmed by multiple TI methods. This is to make sure that the prediction is not biased due to the given parameter setting or the particular algorithm underlying a TI method. The value of using different methods is further supported by our analysis indicating substantial complementarity between the different methods. Second, even if the expected topology is known, it can be beneficial to also try out methods that make less assumptions about the trajectory topology. When the expected topology is confirmed using such a method, it provides additional evidence to the user. When a more complex topology is produced, this could indicate that the underlying biology is much more complex than anticipated by the user.

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

\begin{figure}[htb!]
	\centering\includegraphics[width=\linewidth]{fig/figure_6.pdf}
	\caption{
		\textbf{Demonstration of how a common framework for TI methods facilitates broad applicability using some example datasets.}
		Trajectories inferred by each method were projected to a common dimensionality reduction using multidimensional scaling. For each dataset, we also calculated a ‘consensus’ prediction, by calculating the cordist between each pair of models and picking the model with the highest score on average. \textbf{a}, The top methods applied on a dataset containing a linear trajectory of differentiation dendritic cells, going from MDP, CDP to PreDC. \textbf{b}, The top methods applied on a dataset containing a bifurcating trajectory of reprogrammed fibroblasts. \textbf{c}, A synthetic dataset generated by dyntoy, containing four disconnected trajectories. \textbf{d}, A synthetic dataset generated by dyngen, containing a cyclic trajectory.
	}
	\label{fig:figure_6}
\end{figure}

Our study indicates that the field of trajectory inference is maturing, primarily for linear and bifurcating trajectories (Figure \ref{fig:figure_6}a,b). However, we also highlight several ongoing challenges, which should be addressed before TI can be a reliable tool for analyzing single-cell omics datasets with complex trajectories. Foremost, new methods should focus on improving the unbiased inference of tree, cyclic graph and disconnected topologies, as we found that methods repeatedly overestimate or underestimate the complexity of the underlying topology, even if the trajectory could easily be identified using a dimensionality reduction method (Figure \ref{fig:figure_6}c,d). Furthermore, higher standards for code assurance and documentation could help in adopting these tools across the single-cell omics field. Finally, new tools should be designed to scale well with the increasing number of cells and features. We found that only a handful of current methods can handle datasets with more than 10,000 cells within a reasonable time frame. To support the development of these new tools, we provide a series of vignettes on how to wrap and evaluate a method on the different measures proposed in this study at \href{https://benchmark.dynverse.org}{https://benchmark.dynverse.org}.

We found that the performance of a method can be very variable between datasets, and therefore included a large set of both real and synthetic data within our evaluation, leading to a robust overall ranking of the different methods. However, ‘good-yet-not-the-best’ methods \cite{norel_selfassessmenttrapcan_2011} can still provide a very valuable contribution to the field, especially if they make use of novel algorithms, return a more scalable solution or provide a unique insight in specific use cases. This is also supported by our analysis of method complementarity. Some examples for the latter include PhenoPath, which can include additional covariates in its model, ouija, which returns a measure of uncertainty of each cell’s position within the trajectory, and StemID, which can infer the directionality of edges within the trajectory.

\section{Methods}

\subsection{Trajectory inference methods}

We gathered a list of 71 trajectory inference tools (Supplementary Table 1) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{https://github.com/seandavi/awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{https://github.com/agitter/single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}. We welcome any contributions by creating an issue at https://methods.dynverse.org.

Methods were excluded from the evaluation based on several criteria: (1) not freely available; (2) no code available; (3) superseded by another method; (4) requires data types other than expression; (5) no programming interface; (6) unresolved errors during wrapping; (7) too slow (requires more than 1 h on a 100 $\times$ 100 dataset); (8) does not return an ordering; and (9) requires additional user input during the algorithm (other than prior information). The discussions on why these methods were excluded can be found at \href{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}. In the end, we included 45 methods in the evaluation.

\subsection{Method wrappers}

>>>>

% BRED
